Discovering causal factors with back-to-back regression

### 2019-05-22 paper/paper.tex
small things

\begin{abstract}
Identifying causes from observations is at the core of science. This endeavor
is particularly challenging when i) potential factors are difficult to
manipulate individually and ii) observations are complex and multi-dimensional.
To address this issue, we introduce ``back-to-back regression'' (BFR), a
method designed to identify, from a set of co-varying factors, the factors
that most plausibley account for multidimensional observations. After detailing
the proof of convergence, we show that BFR outperforms least-squares and related
regression techniques, as well as cross-decomposition methods (e.g. canonical
correlation analysis, and partial least squares) on two tasks: causal
identification and out-of-distribution prediction. Finally, we apply BFR to
neuroimaging recordings of 102 subjects reading word sequences. The results
show, as expected, that the early brain responses to words appear to be
specifically caused by low-level visual factors whereas late brain responses
appear to be caused by lexical factors, despite the fact that these factors
covary in the study.
% conclude  open
\end{abstract}

@@

\section{Back an forth regression}

@@

    \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
    \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
    \node[] (reg1) at (5,-3.25) {$\hat{E} = \underbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}} $};
    \end{tikzpicture}
    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$. }
    \label{fig:}
\end{figure}

@@

Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $FEX$ be of rank k (which means F and X are of full rank on the space spanned by the active features of E). Let K be the (nx,k) diagonal matrix built by deleting the rightmost nx-k columns of E. Then, $K=(\frac{I_{k}}{0})$. we may write $E=KK'$, and $FEX=(FK)(K'X)$, with $FK$  and $K'X$ of full rank k.

The rank factorisation property of the pseudo inverse (ref) says that, if $A=BC$, A (m,n), B(m,k) or rank k C(k,n) of rank k, then $A^\dagger=C^\dagger B^\dagger= C' (CC')^{-1}(B'B)^{-1}B'$.

Replacing in the expression of the minimal norm solution yields
\begin{equation}
\begin{aligned}
X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
&=X(K'X)'((K'X)(K'X)')^{-1}((FK)'(FK))^{-1}(FK)' \\
&=(XX')K(K'XX'K)^{-1}(K'F'FK)^{-1} K'F' 
\end{aligned}
\end{equation}

Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{3} & \Sigma_{4}\end{array}\right)$, the top left block being (k,k), since $K=\left(\begin{array}{c}I \\\hline 0\end{array}\right)$, we have $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{3} \Sigma_{1}^{-1}\end{array}\right)$, call S the matrix obtained by adding Nx-k empty columns to this one (or, equivalently multiplying it by K').

>>>>

\begin{abstract}
Identifying causes from observations is at the core of science. This endeavor
is particularly challenging when i) potential factors are difficult to
manipulate individually and ii) observations are complex and multi-dimensional.
To address this issue, we introduce ``back-to-back regression'' (BFR), a
method designed to identify, from a set of co-varying factors, the factors
that most plausible account for multidimensional observations. After detailing
the proof of convergence, we show that BFR outperforms least-squares and related
regression techniques, as well as cross-decomposition methods (e.g. canonical
correlation analysis, and partial least squares) on two tasks: causal
identification and out-of-distribution prediction. Finally, we apply BFR to
neuroimaging recordings of 102 subjects reading word sequences. The results
show, as expected, that the early brain responses to words appear to be
specifically caused by low-level visual factors whereas late brain responses
appear to be caused by lexical factors, despite the fact that these factors
covary in the study.
% conclude  open
\end{abstract}

@@

\section{Back to back regression}

@@

    \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
    \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
    \node[] (reg1) at (5,-3.25) {$\hat{E} = \underbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}} $};
    \end{tikzpicture}
    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$}
    \label{fig:}
\end{figure}

@@

Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $FEX$ be of rank k (which means F and X have full rank on the space spanned by the active features of E). Let K be the $\mathbb{R}^{nx\times k}$ diagonal matrix obtained by deleting the rightmost nx-k columns of E. Then, $E=KK'$, and $FEX=(FK)(K'X)$, with $FK$  and $K'X$ of full rank k.

The rank factorisation property of the pseudo inverse (ref) says that, if $A=BC$ with $A\in\mathbb{R}^{m\times n}$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$ of rank k, then $A^\dagger=C^\dagger B^\dagger$ and since C is full rank, $C^\dagger=C' (CC')^{-1}$.

Replacing in the expression of the minimal norm solution yields
\begin{equation}
\begin{aligned}
X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
&=X(K'X)'((K'X)(K'X)')^{-1}(FK)^\dagger \\
&=(XX')K(K'XX'K)^{-1}(FK)^\dagger 
\end{aligned}
\end{equation}

Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$. Then, $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{2} \Sigma_{1}^{-1}\end{array}\right)$, call S the matrix obtained by adding Nx-k empty columns to this one (or, equivalently multiplying it by K').
